---
title: "MH3511 Project Analysis"
output: html_document
date: "2024-03-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction 

```{r}
# Load required libraries 
library(dplyr)
library(readr)
library(e1071) #to check skewness of data
library(corrplot) #for correlation plot
library(fastDummies) #to construct dummy variables in Statistical Analysis

# Load the dataset
insurance = read.csv("insurance.csv")
```


# 2. Data Description
```{R}
# Removing missing values 
insurance = na.omit(insurance)

# convert children into a categorical variable 
insurance$children = as.factor(insurance$children)

str(insurance) 
```


# 3. Description and Cleaning of dataset

## 3.1 Summary Statistics on the main variable of interest, charges

```{r}
hist(insurance$charges, main="Histogram of Insurance Charges", xlab="Charges")
boxplot(insurance$charges, main="Boxplot of Insurance Charges", ylab="Charges")
skewness(insurance$charges) #1.512483

# Log-transformation: 
hist(log(insurance$charges), main="Histogram of log_charges", xlab="log_charges")
boxplot(log(insurance$charges), main="Boxplot of log_charges", ylab="log_charges")
skewness(log(insurance$charges)) #-0.08989561
summary(log(insurance$charges))
```

Replacing charges with log_charges:
```{r}
insurance = insurance %>%
  mutate(log_charges = log(charges)) %>%
  select(-charges)
```


## 3.2 Summary Statistics 

### 3.2.1 Age

```{r}
hist(insurance$age, main="Histogram of Age", xlab="Age", ylim = c(0,200))
boxplot(insurance$age, main="Boxplot of Age", ylab="Age")
```

### 3.3.2 BMI
```{r}
hist(insurance$bmi, main="Histogram of BMI", xlab="BMI")
boxplot(insurance$bmi, main="Boxplot of BMI", ylab="BMI")

skewness(insurance$bmi) #0.2834106
```

Removing outliers:
```{r}
sum(insurance$bmi > 50) # 3 outliers with bmi > 50 

# Remove the 3 outliers with bmi > 50  
insurance = insurance %>%
  filter(bmi <= 50)

# Data visualisation w/o the 3 outliers
hist(insurance$bmi, main="Histogram of BMI", xlab="BMI")
boxplot(insurance$bmi, main="Boxplot of BMI", ylab="BMI")

skewness(insurance$bmi) #0.2177818
```

### 3.2.3 Children
```{r}
barplot(table(insurance$children), main="Barplot of Number of Children", ylab="Count", ylim = c(0,600))

summary(insurance$children) #Check frequency 

# Recategorise children categories to obtain similar sample size in each group
insurance <- insurance %>%
  mutate(new_children = case_when(
    children == "0" ~ "0",
    children == "1" ~ "1",
    children %in% c("2", "3", "4", "5") ~ "2 or more"
  )) %>%
  select(-children)

insurance$new_children = as.factor(insurance$new_children)

# New barplot with recategorised children groups
barplot(table(insurance$new_children), main="Barplot of Number of Children", ylab="Count", ylim = c(0,600))
```

### 3.2.4 Sex
```{r}
insurance$sex = as.factor(insurance$sex)
barplot(table(insurance$sex), main="Barplot of Sex", ylab="Count", ylim = c(0,700))

summary(insurance$sex) #Frequency for each gender
```

### 3.2.5 Smoker
```{R}
insurance$smoker = as.factor(insurance$smoker)
barplot(table(insurance$smoker), main="Barplot of Smoker", ylab="Count", ylim = c(0,1200))
```

### 3.2.6 Region
```{r}
insurance$region = as.factor(insurance$region)
barplot(table(insurance$region), main="Region", ylab="Count", ylim = c(0,400))
```


## 3.3 Final dataset for Analysis 
```{r}
str(insurance)
```


# 4. Statistical Analysis 

## 4.1 Correlations between log_charges and other continuous variable 
```{r}
numvars = sapply(insurance, is.numeric)

corrplot(cor(insurance[, numvars]), method = "shade", type = "upper", 
         addCoef.col = "black", insig = "p-value", number.cex = 1, tl.cex = 1)
```

## 4.2 Relation between log_charges and age 
```{r}
# Group the ages into 3 groups 
insurance$age_groups <- factor(ifelse(insurance$age < 35, "Youth",
      ifelse(insurance$age < 50, "Middle Aged", "Seniors")),
      levels = c("Youth", "Middle Aged", "Seniors"))

# Boxplots to see spread of distributions between different age groups 
boxplot(log_charges ~ age_groups, data=insurance, main="Boxplot of log_charges against Age Groups")

# Residual Plot and Q-Q Plot
par(mfrow=c(1,2))
plot(lm(log_charges ~ age_groups, insurance), 1:2)

# conduct Kruskal-Wallies Test
kruskal.test(insurance$log_charges, insurance$age_groups)
pairwise.wilcox.test(insurance$log_charges, insurance$age_groups, p.adjust.method = "none")
```

## 4.3 Relation between log_charges and bmi 
```{r}
# Group bmi into obese and not obese
insurance$obesity = factor(
  ifelse(insurance$bmi < 30, "Not Obese", "Obese"),
  levels = c("Not Obese", "Obese")
)

# Boxplots to see spread of distributions between obesity
boxplot(insurance$log_charges ~ insurance$obesity, main = "Boxplot of log_charges based on Obesity", xlab = "Obesity", ylab = "log_charges")

# F-test to check if variances of obesity are the same
var.test(insurance$log_charges ~ insurance$obesity)

# t-test to check if the means of obesity are the same
t.test(insurance$log_charges ~ insurance$obesity, var.equal = F)
```

## 4.4 Relation between log_charges and children
```{r}
# Boxplots to see spread of distributions in different children groups
boxplot(insurance$log_charges ~ insurance$new_children, main = "Boxplot of log_charges based on the Number of Children", xlab = "Number of Children", ylab = "log_charges")

# ANOVA to check equality of means of different groups of children
summary(aov(insurance$log_charges ~ insurance$new_children))

# Pairwise comparisons using t-test to determine which group is significant different from the rest
pairwise.t.test(insurance$log_charges, insurance$new_children, p.adjust.method = "none")
```

## 4.5 Relation between log_charges and sex

```{r}
# Boxplots to see spread of distributions between different sex
boxplot(insurance$log_charges ~ insurance$sex, main = "Boxplot of log_charges based on sex", xlab = "Sex", ylab = "log_charges")

# F-test to check if variances of the different sexes are the same
var.test(insurance$log_charges ~ insurance$sex)

# t-test to check if the means of the different sexes are the same
t.test(insurance$log_charges ~ insurance$sex, var.equal = F)
```

## 4.6 Relation between log_charges and smoker

```{r}
# Boxplots to see spread of distributions based on smoking status 
boxplot(insurance$log_charges ~ insurance$smoker, main = "Boxplot of log_charges by Smoking Status", xlab = "Smoker", ylab = "log_charges")

# Create smoker and non-smoker dataset with log_charges 
smokeyes = insurance %>%
  filter(smoker == "yes") 

smokeno = insurance %>%
  filter(smoker == "no")

# F-test to check if variances of the different smoking status are the same
var.test(smokeyes$log_charges, smokeno$log_charges)

# t-test to check if the means of the different smoking status are the same
t.test(smokeyes$log_charges, smokeno$log_charges, var.equal = FALSE)
```

## 4.7 Relation between log_charges and region

```{r}
# Boxplots to see spread of distributions between different regions 
boxplot(log_charges ~ region, data = insurance, main = "Boxplot of log_charges based on Regions in the U.S.")

# ANOVA to check equality of means of different regions
region_aov <- aov(log_charges ~ region, data=insurance)
summary(region_aov)
```


## 4.8 Testing for Association between Significant Categorical Variabless

### 4.8.1 Chi-Square Test for age_groups and obesity
```{r}
# Construct 2-way Contingency Table 
age_groups_obesity = with(insurance, table(age_groups, obesity))
age_groups_obesity

# Conduct Chi-square test
chisq.test(age_groups_obesity) # p-value = 0.0008557

# Interaction Diagram
interaction.plot(insurance$age_groups, insurance$obesity, insurance$log_charges, xlab="Age Groups", ylab="log_charges", trace.label = "Obesity", col=1:4)

```

### 4.8.2 Chi-Square Test for age_groups and new_children
```{r}
# Construct 2-way Contingency Table
age_groups_new_children = with(insurance, table(age_groups, new_children))
age_groups_new_children

# Conduct Chi-square test
chisq.test(age_groups_new_children) # p-value < 2.2e-16

#Interaction Diagram
interaction.plot(insurance$new_children, insurance$age_groups, insurance$log_charges, xlab="Number of Children", ylab="log_charges", trace.label = "Age Groups", col=1:3)
```

### 4.8.3 Chi-Square Test for age_groups and smoker
```{r}
# Construct 2-way Contingency table 
age_groups_smoker = with(insurance, table(age_groups, smoker))
age_groups_smoker

# Conduct Chi-square test
chisq.test(age_groups_smoker)
```

### 4.8.4 Chi-Square Test for obesity and new_children
```{r}
# Construct 2-way Contingency Table 
obesity_new_children = with(insurance, table(obesity, new_children))
obesity_new_children

# Conduct Chi-square test
chisq.test(obesity_new_children) # p-value = 0.3507

```

### 4.8.5 Chi-Square Test for obesity and smoker
```{r}
# Construct 2-way Contingency Table 
obesity_smoker = with(insurance, table(obesity, smoker))
obesity_smoker

# Conduct Chi-square test
chisq.test(obesity_smoker) # p-value = 1

```

### 4.8.6 Chi-Square Test for new_children and smoker
```{r}
new_children_smoker = with(insurance, table(new_children, smoker))
new_children_smoker
chisq.test(new_children_smoker) # p-value = 0.4487
```

## 4.9 Multiple Linear Regression

```{r}
# Initial Model 
lm_model <- lm(log_charges ~ age + bmi + new_children + smoker + sex + region, data = insurance)
summary(lm_model) # r squared = 0.7686
```

```{R}
# Create dummy variables 
insurance <- dummy_cols(insurance, select_columns = "new_children")
insurance <- dummy_cols(insurance, select_columns = "smoker")

# Interaction 1: bmi * age 
insurance$interaction1 <- insurance$bmi * insurance$age

# Interaction 2: age * new_children1
insurance$interaction2 <- insurance$age * insurance$new_children_1

# Interaction 3: age * new_children 2 or more
insurance$interaction3 <- insurance$age * insurance$'new_children_2 or more'

# Model 1 with Interaction 1
model_interaction1 <- lm(log_charges ~ age + bmi + new_children + smoker_yes + sex + region + interaction1, data = insurance) 
summary(model_interaction1) #r^2 = 0.7691, p > 0.05

# Model 2 with Interaction 2
model_interaction2 <- lm(log_charges ~ age + bmi + new_children + smoker_yes + sex + region + interaction2, data = insurance)
summary(model_interaction2) #r^2 = 0.7686, p > 0.05

# Model 3 with Interaction 3
model_interaction3 <- lm(log_charges ~ age + bmi + new_children + smoker_yes + sex + region + interaction3, data = insurance) 
summary(model_interaction3) # r^2 = 0.7749, p = 1.42e-09

# Model 4 with Interactions 2,3
model_interaction2_and_3 <- lm(log_charges ~ age + bmi + new_children + smoker_yes + sex + region + + interaction2 + interaction3, data = insurance) 
summary(model_interaction2_and_3)
# r^2 = 0.7752 (HIGHEST)
# BUT P VALUE OF INTERACTION2 NOT SIGNIF = 0.182904

# Model 5 with Interactions 1,2,3
model_interaction1_2_and_3 <- lm(log_charges ~ age + bmi + new_children + smoker_yes + sex + region + interaction1 + interaction2 + interaction3, data = insurance) 
summary(model_interaction1_2_and_3)
# r^2: 0.7756
# p value of int1 and int2 not signif 
# p value for int3 = 8.04e-10
```

Without sex and region: 
```{r}
model_initial_revised <- lm(log_charges ~ age + bmi + new_children + smoker_yes, data = insurance) 
summary(model_initial_revised) # r2= 0.7629 # p < 0.05 for all variables

#### MODEL WITH INTERACTION 1 ONLY ####
model_interaction1_revised <- lm(log_charges ~ age + bmi + new_children + smoker_yes + interaction1, data = insurance) #old r^2 = 0.7691 # p > 0.05
summary(model_interaction1_revised) # new r^2: 0.7635 #  p = 0.089705

#### MODEL WITH INTERACTION 2 ONLY ####
model_interaction2_revised <- lm(log_charges ~ age + bmi + new_children + smoker_yes + interaction2, data = insurance)
summary(model_interaction2_revised)
# new r^2 = 0.763 # 0.737

#### MODEL WITH INTERACTION 3 ONLY ####
model_interaction3_revised <- lm(log_charges ~ age + bmi + new_children + smoker_yes  + interaction3, data = insurance) 
summary(model_interaction3_revised) #r^2: 0.7691 # p = 3.03e-09 for int3

### MODEL WITH INTERACTION 2&3 ###
### I.E. INTERACTION EFFECTS BETWEEN AGE AND CHILDREN ###
model_interaction2_and_3_revised <- lm(log_charges ~ age + bmi + new_children + smoker_yes  + interaction2 + interaction3, data = insurance) 
summary(model_interaction2_and_3_revised)
# r^2 = 0.7694 
# BUT P VALUE OF INTERACTION2 NOT SIGNIF = 0.21556


### MODEL WITH INTERACTIONS 1,2,3 ###
model_interaction1_2_and_3_revised <- lm(log_charges ~ age + bmi + new_children + smoker_yes + interaction1 + interaction2 + interaction3, data = insurance) 
summary(model_interaction1_2_and_3_revised) #r^2: 0.7698
# p = 0.123897, p= 0.199192
```


